Debugging by \textit{lastChange}

### 2010-10-22 papers/pldi2011/sigplanconf-template.tex
combine two subsections: conditional breakpoints and reproducible non-determinism

\section{Discussion}
\subsection{Converting \textit{lastChange} result to a conditional breakpoint}
We believe that in many cases it is possible to convert the
\textit{lastChange} result to a conditional breakpoint and therefore
debugger is able to pause at the point instead of showing collected
data to the developer. Here, we present an example which shows a case
this transformation is not possible.

Figure ~\ref{fig:counter-example} shows a function with processes an
array. The array contains numbers except one itme which is
\texttt{undefined} and it causes a bug in line 30. There is a call to
\texttt{randomPermutation} function in line 11 which randomly
permutates the array item. So at every execution the
\texttt{undefined} item will be in a new place. Calling
\textit{lastChange} on \texttt{x} at the place bug happens, gives a
point which shows line 14. Although this point exists at every
re-execution but it can not be identified by a conditional breakpoint.

@@

%---------------------------------------------------------------------------------------------------
\subsection{Reproducible Non-deterministic Execution}
Thus far we have not discussed problems caused by mul-tiple threads or
other sources of non-deterministic execu-tions. We want to explain why
we believe \textit{lastChange} is robust in the practically important
case where a bug is reproducible even though the execution may not be
deterministic.  Because \textit{lastChange} require re-execution, we
rely on reproducible but not necessarily deterministic execution. A
bug is reproducible for a developer when the developer can start from
a determined initial state, operate on the program with a list of
actions, and reproduce the symptoms of the bug. The details of the
execution can change each time we re-execute the buggy program, but
the buggy result is the same. The entire query chain reapplies during
each execution so the data we show the developer will be internally
consistent. The reproducibility of the bug means that the defect is
very unlikely to depend on the order of events during the execution.

In this important case of reproducible bugs, \textit{lastChange} are
more effective than breakpoints. In the case of logically
deterministic program execution, we can use the result from a
\textit{lastChange} operation to set a conditional breakpoint then
re-execute the program to position the execution trace backwards from
our first breakpoint.

%(This may be a useful adjunct for Querypoint debuggers to implement, but this backwards motion in the execution logic is not required for Querypoint debugging.) Thus in this case Querypoint debugging can do the same kinds of things as conventional breakpoints just more automatically.

In the case of a non-deterministic program, a \textit{lastChange} is
not equivalent to any series of conventional watchpoints or
breakpoints. Each time we re-execute a non-deterministic program, the
details of execution in-struction order may change. For example, if we
record the source code lines every time a conventional watchpoint
hits, the record may differ each time we re-execute. Sup-pose we
consult one such record and set a breakpoint on the last entry, the
apparent lastChange source line. When we re-execute, the breakpoint
will hit, but the information we gain may be incorrect: this may not
be the lastChange for this particular re-execution. The
\textit{lastChange} method records the values we need and the sequence
of source lines from all of the watchpoints, then analyzes the record
to select the correct lastChange point. The data shown to the
developer will be internally consistent, but of course it may change
from a previous re-execution, surprising the developer. This is just a
signal that the execution is not deterministic. In future we hope to
compare queries from successive executions as a tool for learning
about non-deterministic executions.

>>>>

\section{Discussion}

%---------------------------------------------------------------------------------------------------
\subsection{Reproducible Non-deterministic Execution}

By construction \textit{lastChange} works whenever conventional
breakpoint debugging works. By that we mean that, since we use
conventional breakpoint debugging technology under the covers,
\textit{lastChange} works whenever breakpoint debugging
works. However, there are cases where breakpoint debugging fails but
\textit{lastChange} works.  These are cases of \textit{reproducible
  non-deterministic} execution.

A bug is \textit{reproducible} for a developer when the developer can
start from a determined initial state, operate on the program with a
list of actions, and reproduce the symptoms of the bug. The details of
the execution can change each time we re-execute the buggy program,
but the buggy result is the same.  The reproducibility of the bug
means that the defect is very unlikely to depend on the order of
events during the execution. When the developer selects a value and
asks for \textit{lastChange}, they are expressing an hyptothesis that
the value is related to the defect. Therefore the reproducibility of
the bug makes it very unlikely that the \textit{lastChange} will
depend upon order of events during execution.

Since \textit{lastChange} does not rely on determinism, some bugs that
confound breakpoint debuggers will be easier to find with
\textit{lastChange}.  Consider the following example.
Figure ~\ref{fig:counter-example} shows a function with processes an
array. The array contains numbers except one itme which is
\texttt{undefined} and it causes a bug in line 30. There is a call to
\texttt{randomPermutation} function in line 11 which randomly
permutates the array item. So at every execution the
\texttt{undefined} item will be in a new place. Calling
\textit{lastChange} on \texttt{x} at the place bug happens, gives a
point which shows line 14. Although this point exists at every
re-execution but it can not be identified by a conditional breakpoint.

@@

Thus in the case of a non-deterministic program, a \textit{lastChange}
is not equivalent to any series of conventional watchpoints or
breakpoints. The extra power of \textit{lastChange} derives from its
analysis of the entire execution up to the reproduction point and its
memory of previous work on the defect.  Each time we re-execute a
non-deterministic program, the details of execution instruction order
may change. For example, if we record the source code lines every time
a conventional watchpoint hits, the record may differ each time we
re-execute. But \textit{lastChange} does not ask the developer to
consider these single points.  Rather we analyze all of the
watchpoints in an entire execution at the reproduction point where we
know the defect has already occured and only show the final result to
the developer.

Of course there can be cases where the defect is reproducible but the
value from \textit{lastChange} is not. As in conventional breakpoint
debugging, we can only determine this empirically, by observing
multiple executions with different values. Unlike conventionaly
breakpoint debugging, implementations of \textit{lastChange} readily
aid in detecting such cases: as we work back through a series of
\textit{lastChange} requests we can compare values with previous
executions. Different values on different executions will signal that
the execution is not deterministic. In future we hope to compare
queries from successive executions as a tool for learning about
non-deterministic executions.
